Topological order in the quantum dimer-pentamer model

### 2015-10-07 qdpm_paper.tex
quick
Started appedix on gap calculations

        In the QDPM Ising degrees of freedom live on the links of the lattice.
        The operator $\hat{\sigma}^x_l$ acts on these links and has 
        eigenvalues $\sigma_l = \pm 1$, where $\sigma_l$ is the
        Ising variables of the $l^{\mathrm{th}}$ link on the lattice. The number operator for a give link is
        defined as $\hat{n}_l \equiv (1+\hat{\sigma}_x)/2$ and counts the number of dimers on link $l$.
        An exactly local gauge symmetry exists on the lattice due to the constraint that either one or
        four dimers touch any give vertex. In terms of $n_l$ we can define an operator that counts the
        number of dimers touching a given vertex $v$ as $n_v = \sum_{l\in v} n_l$.

@@

        We work specifically at the RK point at which the ground state wave function
        is an equal weighted superposition of fully packed dimer pentamer configurations
        %
        \begin{equation}
            \label{}
            \left| \Psi \right\rangle = \frac{1}{\sqrt{N}}\sum_{\{C\}} \left| C \right\rangle
            ,
        \end{equation}
        %
        where $\{C\}$ is the set of fully packed configurations. The Hilbert space can be defined
        using the gauge transformation in Eq.~\ref{eqn:gauge_trans} as all $\left| C \right\rangle$
        satisfying $G \left| C \right\rangle = \left| C \right\rangle$.

        The QDPM Hamiltonian
        \begin{equation}
            H_{\mathrm{QDPM}} = H_{\mathrm{QDM}} + \mathrm{pentamer\ terms}
        \end{equation}

        The pentamer terms in the Hamiltonian can be constructed with the dynamics shown in
        Fig.~\ref{fig:pentamer_moves}.  
        \begin{figure}[htpb]
            \centering
            \includegraphics[width=0.8\linewidth]{pentamer_moves.pdf}
            \caption{a) Horizontal creation an annihilation, b) horizontal move, c) and d) are two
            different possible diagonal moves in the same direction.}
            \label{fig:pentamer_moves}
        \end{figure}
        Each of the  moves shown in Fig.~\ref{fig:pentamer_moves} can be
        rotated to produce the symmetry related counterpart (eg. vertical creation and annihilation). 
        As an example the creation and annihilation term in the Hamiltonian is
        \begin{multline}
        H_{\mathrm{pentamer\ creation}} = t_{\mathrm{create}}
            \sum_{\bstr}
            \left(
                \left|
                    \hprs
                \right\rangle
                \left\langle
                    \hspr
                \right|
                +
                \mathrm{h.c.}
            \right)
            \\
            - v_{\mathrm{create}}
            \sum_{\bstr}
                \left(
                    \left|
                        \hprs
                    \right\rangle
                    \left\langle
                        \hprs
                    \right|
                    +
                    \left|
                        \hspr
                    \right\rangle
                    \left\langle
                        \hspr
                    \right|
                \right)
        \end{multline}
        where the factors $t_{\mathrm{create}}$ and $v_{\mathrm{create}}$ are the coefficients of
        the pentamer pair creation and annihilation respectively. At the RK point
        $t/v=1$ for each all the pentamer terms. 
        Writing the kinetic and potential energy terms as done above for all the dynamics shown in
        Fig.~\ref{fig:pentamer_moves} and all of their symmetry related counterparts comprise the
        the pentamer terms of the Hamiltonian.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Estimating the Gap}

%    \begin{table*}[htpb]
%    \setlength\tabcolsep{0.3cm}
%        \centering
%        \caption{$\Delta \tau = 150$}
%        \label{tab:label}
%        \begin{tabular}{l l l l l}
%            \hline\hline
%         L            		& 8x8                               		& 12x12                            		& 16x16					& 18x18 \\ 
%            \hline                                                                                                                           
%         frac 1.0     	&                                       		&                                  		&                                    		&  \\
%            \hline                                                                                                                           
%         Dimer        	&  $ 3(1) \times 10^{-2}$        	& $ 1.43(3) \times 10^{-2}$	& $0.81(2)\times 10^{-2}$  	& \\
%         $Z_2$ origin 	&  $ 0.787?(5)\times 10^{-2} $ 	& $ 0.344? (1)\times 10^{-2}$ 	& $0.198? (4)\times 10^{-2} $ 	&  \\
%         $Z_3$ origin 	&  NA                                   	& NA                               		& NA                                  	& \\
%            \hline                                                                                                                         
%         frac 0.9     	&                                       		&                                  		&                                     		&  \\
%            \hline                                                                                                                          
%         Dimer origin 	&  $ 0.419(1)\times 10^{-2} $    	& $ 0.406(2)\times 10^{-2} $ 	& $0.429(1)\times 10^{-2} $   	&  $ 0.405(2)\times 10^{-2} $\\
%         $Z_2$ origin 	&  $ 0.216?(2)\times 10^{-2} $ 	& $ 0.141?(1)\times 10^{-2} $ 	& $0.129?(2)\times 10^{-2} $	&  $ 0.126?(2)\times 10^{-2} $\\
%         $Z_3$ origin 	&  $ 0.346?(4)\times 10^{-2} $ 	& $ 0.348(1)\times 10^{-2} $ 	& $0.359?(4)\times 10^{-2} $	&  $ 0.355?(3)\times 10^{-2}$\\
%            \hline                                                                                                                          
%         frac 0.5     	&                                       		&                                   		&                                    		& \\
%            \hline                                                                                                                          
%         Dimer origin 	&                                       		& $ 2.075(9)\times 10^{-2} $ 	& $2.05(1) \times 10^{-2}$		&  $ $ \\
%         $Z_2$ origin 	&                                       		& $ 0.615?(2)\times 10^{-2} $ 	& $0.614?(3)\times 10^{-2} $	&  $ $ \\
%         $Z_3$ origin 	&                                       		& $ 1.939(2)\times 10^{-2} $ 	& $1.927(2)\times 10^{-2} $	&  $ $  
%        \end{tabular}
%    \end{table*}

    \begin{table*}[htpb]
    \setlength\tabcolsep{0.3cm}
        \centering
        \caption{$\Delta \tau = 250$}
        \label{tab:label}
        \begin{tabular}{l l l l}
            \hline\hline
                        & Dimer                 & $Z_2$ double ($L/2$)  &  $Z_3$ single (origin)     \\   
            \hline
            frac 1.0    &  &  & \\
            $8\times8$  & $3.(3)\times10^{-2}$  &                       &                            \\
            $12\times12$& $1.4(5)\times10^{-2}$ &                       &                            \\
            $16\times16$&                       &                       &                            \\
            $18\times18$&                       &                       &                            \\
            \hline
            frac 0.9    &  &  & \\
            $8\times8$  & $0.35(6)\times10^{-2}$&$0.102(1)\times10^{-2}$& $0.306(3)\times10^{-2}$     \\
            $12\times12$& $0.35(7)\times10^{-2}$&$0.25(5)\times10^{-2}$ & $0.316(4)\times10^{-2}$    \\
            $16\times16$& $0.35(7)\times10^{-2}$&$0.36(3)\times10^{-2}$ & $0.316(8)\times10^{-2}$    \\
            $18\times18$&                       &                       &                            \\
            \hline
            frac 0.5    &  &  & \\
            $8\times8$  & $1.0(6)\times10^{-2}$ &$0.05(5)\times10^{-2}$ & $1.7(1)\times10^{-2}$     \\
            $12\times12$& $1.5(7)\times10^{-2}$ &                       & $1.9(7)\times10^{-2}$      \\
            $16\times16$&                       &                       &                            \\
            $18\times18$&                       &                       &                            \\
        \end{tabular}
    \end{table*}

@@

\bibliographystyle{apsrev4-1}
\bibliography{QDPM}

>>>>

        In the QDPM Ising degrees of freedom live on the links of the lattice.
        The operator $\hat{\sigma}^x_l$ acts on these links with 
        eigenvalues $\sigma_l = \pm 1$, where $\sigma_l$ is the
        Ising variables of the $l^{\mathrm{th}}$ link on the lattice. The number operator for a give link is
        defined as $\hat{n}_l \equiv (1+\hat{\sigma}_x)/2$ and counts the number of dimers on link $l$.
        An exactly local gauge symmetry exists on the lattice due to the constraint that either one or
        four dimers touch any give vertex. In terms of $n_l$ we can define an operator that counts the
        number of dimers touching a given vertex $v$ as $n_v = \sum_{l\in v} n_l$.

@@

        We work specifically at the RK point where the ground state wave function
        is an equal weighted superposition of fully packed dimer pentamer configurations
        %
        \begin{equation}
            \label{}
            \left| \Psi \right\rangle = \frac{1}{\sqrt{N}}\sum_{\{C\}} \left| C \right\rangle
            ,
        \end{equation}
        %
        where $\{C\}$ is the set of fully packed configurations. The Hilbert space can be defined
        using the gauge transformation in Eq.~\ref{eqn:gauge_trans} as all $\left| C \right\rangle$
        satisfying $G \left| C \right\rangle = \left| C \right\rangle$.

        The QDPM Hamiltonian comprises the QDM Hamiltonian pluss additional terms coresponding to
        the pentamer dynamics.
        \begin{equation}
            H_{\mathrm{QDPM}} = H_{\mathrm{QDM}} + \mathrm{pentamer\ terms}
        \end{equation}
        Specificaly the penatamers pentamer terms are constructed with the dynamics shown in
        Fig.~\ref{fig:pentamer_moves}.  
        \begin{figure}[htpb]
            \centering
            \includegraphics[width=0.8\linewidth]{pentamer_moves.pdf}
            \caption{a) Horizontal creation an annihilation, b) horizontal move, c) and d) are two
            different possible diagonal moves in the same direction.}
            \label{fig:pentamer_moves}
        \end{figure}
        Each of the  moves shown in Fig.~\ref{fig:pentamer_moves} can be
        rotated to produce the symmetry related counterpart (eg. vertical creation and annihilation). 
        As an example the creation and annihilation term in the Hamiltonian is
        \begin{multline}
        H_{\mathrm{pentamer\ creation}} = t_{\mathrm{create}}
            \sum_{\bstr}
            \left(
                \left|
                    \hprs
                \right\rangle
                \left\langle
                    \hspr
                \right|
                +
                \mathrm{h.c.}
            \right)
            \\
            - v_{\mathrm{create}}
            \sum_{\bstr}
                \left(
                    \left|
                        \hprs
                    \right\rangle
                    \left\langle
                        \hprs
                    \right|
                    +
                    \left|
                        \hspr
                    \right\rangle
                    \left\langle
                        \hspr
                    \right|
                \right)
        \end{multline}
        where the factors $t_{\mathrm{create}}$ and $v_{\mathrm{create}}$ are the coefficients of
        the pentamer pair creation and annihilation respectively. At the RK point
        $t/v=1$ for each all the pentamer terms. 
        Writing the kinetic and potential energy terms as done above for all the dynamics shown in
        Fig.~\ref{fig:pentamer_moves} and all of their symmetry related counterparts comprise the
        the pentamer terms of the Hamiltonian.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Estimating the Gap}
    An exponetial decay of the spacial correlations shown in the previous sections 
    suggeste a gap in the spectrum. However we would like to know if the lowest lying excitations are in the vison
    or the dimer sector.  The energy gap of the dimer and vison exsitations can be estimated by investigating the dimer and
    vison imaginary time correlations respectivly as outlined by Henley \cite{Henley2004a}. 
    One changlenge in the gap estimates from the imaginary time correlations is going far enough out
    in imaginary time. In appendix \ref{appendix_gap_err} we discuss our proccess in determining
    that this systematic error is sufficinetly small which is necissary when compairing gap
    estimates for different excitatoins.

    %At the RK point the Hamiltonian is proportional to the matrix of transitions rates the eigen modes of wich describe the time evolution of the
    %system. Since the Hamiltonian and the rate matrix share eigenvalues the lowest energy can be
    %found by looking at the fit of the imaginary time correlations far out in imaginary time where
    %the smalles exponetial dominates.

    In this paper we only investigate the high symetry points in the Brillouin zone.

    In Table~\ref{tab:gap_table} we show the estimates of the dimer gap, the double $Z_2$ vison gap
    at $L/2$ and the $Z_3$ vison gap.
    \begin{table*}[htpb]
    \setlength\tabcolsep{0.3cm}
        \centering
        \caption{$\Delta \tau = 250$}
        \label{tab:gap_table}
        \begin{tabular}{l l l l}
            \hline\hline
                        & Dimer                 & $Z_2$ double ($L/2$)  &  $Z_3$ single (origin)     \\   
            \hline
            frac 1.0    &  &  & \\
            $8\times8$  & $3.24(5)\times10^{-2}$  &                       &                            \\
            $12\times12$& $1.47(5)\times10^{-2}$ &                       &                            \\
            $16\times16$&                       &                       &                            \\
            $18\times18$&                       &                       &                            \\
            \hline
            frac 0.9    &  &  & \\
            $8\times8$  & $0.3549(9)\times10^{-2}$&$0.0993(3)\times10^{-2}$& $0.3069(3)\times10^{-2}$     \\
            $12\times12$& $0.357(2)\times10^{-2} $&$0.256(2)\times10^{-2} $ & $0.3164(7)\times10^{-2}$    \\
            $16\times16$& $0.357(3)\times10^{-2} $&$0.363(5)\times10^{-2} $ & $0.3170(9)\times10^{-2}$    \\
            $18\times18$& $0.352(2)\times10^{-2} $&$0.419(3)\times10^{-2} $ & $0.3167(7)\times10^{-2}$    \\
            \hline
            frac 0.5    &  &  & \\
            $8\times8$  & $1.0(6)\times10^{-2}$ &$0.05(5)\times10^{-2}$ & $1.7(1)\times10^{-2}$     \\
            $12\times12$& $1.5(7)\times10^{-2}$ &                       & $1.9(7)\times10^{-2}$      \\
            $16\times16$&                       &                       &                            \\
            $18\times18$&                       &                       &                            \\
        \end{tabular}
    \end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Error in the Gap Estimations}
\label{appendix_gap_err}

When estimating the gap by calculating correlations in imaginary time one has to be sure
that the range of imaginary time is appropriate otherwise it is not meaningful to compare the relative 
sizes of the vison and dimer gaps. In
%Fig.~\ref{fig:HERE} 
we show the finite size scaling of the different gaps. For each gap estimation we wanted to reduce
if not eliminate the systematic error due to and inadiquate range of imaginary time. This was
accived by investegating the gap as a fuction of the lower cutoff in the exponential fit of. We
extended the range of $\tau$ until the gap plataued and had little dependence on the cutoff. We show an
example of the estimate of the $Z_3$ vison gap dependence on the lower cutoff of the fit for the
$HERE\times HERE$ lattice with a fraction (HERE) of $0.9$ in 
%Fig.~\ref{fig:}
%\begin{figure}[htpb]
%    \centering
%    \includegraphics[width=0.8\linewidth]{name.ext}
%    \caption{Name}
%    \label{fig:name}
%\end{figure}

NOTE: There is also some systematic error in the discritization of time. This error falls off in the QDM
as the number of flippable plaquestts. We assume this error it is small compared to the
statistical error.

\section{Imaginary time Vison Correlations}
\label{appendix_tau_corr}

    \subsection{$Z_2$ Single and Double Visons}

        \subsubsection{Single Vison}
            First we distiguish between the single and double vison correlation fuctions. We start with the
            the described by Ivanov \cite{Ivanov2004} for rewriting the single vison correlation function in a
            convieient form to implement numericaly. 
            The correlation function between two point visons $V(\bm{x}_1,\tau_1)$, and $V(\bm{x}_2,\tau_2)$ at 
            points in space $\bm{x}_1$, $\bm{x}_2$ and in imaginary time $\tau_1$ and $\tau_2$ can be writen as
            \begin{equation}
                C_{\mathrm{single}}(\bm{x}_2-\bm{x}_1,\tau_2 - \tau_1) = \langle V(\bm{x}_1,\tau_1) V(\bm{x}_2,\tau_2) \rangle
            \end{equation}
            For the $Z_2$ vison $V(\bm{x}_i,\tau_j)^2=1$ and inserting this for $i=2$ and $j=1$ between the two vison
            operators in the above equation we get 
            \begin{equation}
                V(\bm{x}_1,\tau_1) V(\bm{x}_2,\tau_2) = \big(V(\bm{x}_1,\tau_1) V(\bm{x}_2,\tau_1)\big)
                \big( V(\bm{x}_2,\tau_1) V(\bm{x}_2,\tau_2) \big),
                \label{eqn:expanded_z2_v_tau_cor}
            \end{equation}

            The first product in Eq.~\ref{eqn:expanded_z2_v_tau_cor} on the RHS is a 
            string operator between two points on the lattice at a given imaginary time $\tau_2$ 
            \begin{equation}
                V(\bm{x}_1,\tau_1) V(\bm{x}_2,\tau_1) = (-1)^{n},
                \label{eqn:simp_z2_vison_phase}
            \end{equation}
            $n$ being the number of dimers intersecting the path between $\bm{x}_1$ and $\bm{x}_1$. This number
            can not depend on the path so one must choose a meathod of ``gauge fixing''. This can be done
            by choosing an imutable bacround configureation which also changes the phase of the
            operator $V_{12}(\tau)$ in the same way as the dynamic dimers. 
            Including the gauge fixing Eqn.~\ref{eqn:simp_z2_vison_phase} becomes
            \begin{equation}
                V(\bm{x}_1,\tau_1) V(\bm{x}_2,\tau_1) = (-1)^{n+n_{\mathrm{backround}}}.
            \end{equation}

            The second product in Eq.~\ref{eqn:expanded_z2_v_tau_cor} on the RHS is a string operator 
            between two points at the location but at different imaginary time $\tau_2$. 
            In order for the full path in space and
            imaginary time to maintain its path independece it can be shown that the phase 
            changes with every plaquet flip at $\bm{x}_2$. So
            \begin{equation}
                V(\bm{x}_2,\tau_1) V(\bm{x}_2,\tau_2) = (-1)^{n_{\mathrm{flipped}}},
            \end{equation}
            where $n_{\mathrm{flipped}}$ is the number of times the plaquette at $\bm{x}_2$ is flipped.
            Finaly Eq.~\ref{eqn:expanded_z2_v_tau_cor} can be expresed as 
            \begin{equation}
                V(\bm{x}_1,\tau_1) V(\bm{x}_2,\tau_2) =
                (-1)^{n+n_{\mathrm{backround}+n_{\mathrm{flipped}}}}.
            \end{equation}
            The gap that is being estimated by looking at the single vison correlatoin function is
            the energy gap of a single vison exciation, only physical when the system has
            boundries but measureable in any system topology.

        \subsubsection{Double Vison}
            Calculating the doublue vison imaginary time correlation funciton
            \begin{equation}
                C_{\mathrm{double}}(\bm{x}_{12},\tau_2 - \tau_1) = \langle V(\bm{x}_1,\tau_1)V(\bm{x}_2,\tau_1) V(\bm{x}_1,\tau_2)V(\bm{x}_2,\tau_2) \rangle
            \end{equation}
            where the first product on the RHS is the string operator acting along path between points
            $\bm{x}_1$ and $\bm{x}_1$ at imaginary time $\tau_1$ and the second product is the
            string operator along a path between the same points but at $\tau_2$.

            by looking at the double vison correlation function in imaginary time the gap of a pair
            of visons is extracted.

    \subsection{$Z_3$ Single and Double Visons}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{apsrev4-1}
\bibliography{QDPM}
